---
title: "Denis Ostroushko - PUBH 7440 - HW1"
format: pdf
execute: 
  warning: false
  message: false 
  echo: false 
---

```{r}
library(tidyverse)

options(scipen = 99999, 
        knitr.kable.NA = '')
```

# Problem 1



# Problem 2

### 2 - A 

According to given parameters, prior distribution of $\theta$ is given by $\Large \theta$ ~ $\Large N(0, 2)$

And the distribution of data, random variable $Y$ is also normal, $\Large y| \theta$ ~  $\Large N(\theta, 2)$

Using given derivation of posterior distribution, we obtain the distribution of $\theta$ given the observed values of data $y$. 

Going forward, we fix $y$ = 4, we let $\large B = \frac{2}{2+2} = \frac{1}{2}$. 
$\Large \theta|y$ ~ $\Large N(B \times \mu + (1-B) \times y, (1-B) \times \sigma^2) = N(\frac{1}{2} \times 0 + \frac{1}{2} \times 4, \frac{1}{2} \times 2) = N(2, 1)$. 

So, $\Large \theta|y$ ~ $N(2, 1)$

To plot likelihood of observing $y$= 4, I will use the distribution of data $y|\theta$, I will vary the value of $\theta$ and keep variance of $Y$ fixed at 2. 
Since $\theta$ is centered at 0 with variance 2, I will obtain values that correspond to prior distribution of $\theta$

```{r prior likelihood and posterior plot for 2-a}

# prior density save data 

prior_x = seq(from = 0 - 2*3, to = 0+2*3, by = 0.1)
dnorm(
  x = prior_x, 
  mean = 0, 
  sd = sqrt(2)
) -> prior_plot_data 

# posterior 
post_x = seq(from = 2 - 3*1, to = 2+3*1, by = 0.1)
dnorm(
  x = post_x, 
  mean = 2, 
  sd = sqrt(1)
) -> posterior_plot_data 

# likelihood of Y given different parameters of Theta 
y = 4 
means = seq(from = 4 - 2*3, to = 4+2*3, by = 0.1)
likelihood_y = dnorm(x = y, mean = means, sd = sqrt(2))

# plotting data 

plot_df <- 
  data.frame(
    x = c(prior_x, post_x, means), 
    y = c(prior_plot_data, posterior_plot_data, likelihood_y), 
    type = c(rep("Prior of Theta", length(prior_x)), 
             rep("Posterior of Theta", length(post_x)), 
             rep("Likelihood of Y=4", length(likelihood_y)))
  )

ggplot(data = plot_df, 
       aes(x = x, y = y, color = type, group = type)) + 
  theme_minimal() + 
  geom_line() + 
  labs(x = expression(paste("Values of  ", theta)), 
       y = "Density", 
       color = "Curve Type"
  ) 

```

### 2 - B 

When $\tau^2$ = 18, $\large B = \frac{2}{2  + 18} = \frac{1}{10}$. The resulting distributions are then 

$\Large \theta$ ~ $N(0, 18)$

$\Large y|\theta$ ~ $N(\theta, 2)$

Assuming $y$= 4, $\Large \theta|y$ ~ $N(0.1 * 0 + 0.9 * y, 0.9 * 2) = N(3.6, 1.8)$

Prior, marginal likelihood, and posterior distributions take on these shapes: 

```{r prior likelihood and posterior plot for 2-b}

# prior density save data 

prior_x = seq(from = 0 - sqrt(18)*3, to = 0+sqrt(18)*3, by = 0.1)
dnorm(
  x = prior_x, 
  mean = 0, 
  sd = sqrt(18)
) -> prior_plot_data 

# posterior 
post_x = seq(from = 3.6 - 10*sqrt(1.8), to = 3.6 + 10*sqrt(1.8), by = 0.1)
dnorm(
  x = post_x, 
  mean = 3.6, 
  sd = sqrt(1.8)
) -> posterior_plot_data 

# likelihood of Y given different parameters of Theta 
y = 4 
means = seq(from = 4 - 2*3, to = 4+2*3, by = 0.1)
likelihood_y = dnorm(x = y, mean = means, sd = sqrt(2))

# plotting data 

plot_df <- 
  data.frame(
    x = c(prior_x, post_x, means), 
    y = c(prior_plot_data, posterior_plot_data, likelihood_y), 
    type = c(rep("Prior of Theta", length(prior_x)), 
             rep("Posterior of Theta", length(post_x)), 
             rep("Likelihood of Y=4", length(likelihood_y)))
  )

ggplot(data = plot_df, 
       aes(x = x, y = y, color = type, group = type)) + 
  theme_minimal() + 
  geom_line() + 
  labs(x = expression(paste("Values of ", theta)), 
       y = "Density", 
       color = "Curve Type"
  ) 

```

# Problem 3 

### 3 - A 

The chance from picking a ball from bucket 2 is governed by a fair six-sided die. We pick from bucket 2 when we roll 5 or 6, so,
$P(Draw \ from \ 2 ) = \frac{1}{3}$

### 3 - B 

Probability of drawing a blue ball is a weighted average of probabilities that correspond to the bucket that we draw from. 
$P(Blue) = P(Draw \ from \ 1) \times P(Blue|Draw \ from \ 1) + P(Draw \ from \ 2) \times P(Blue|Draw \ from \ 2) = \frac{2}{3} \times \frac{17}{17 + 35} + \frac{1}{3} \times \frac{37}{37 + 23}$ = `r round(2/3 * 17/(17+35) + 1/3 * 37/(37+23), 2)`

### 3 - C 

We need to use Bayes Rule to find this probability from available data 

$P(Draw \ from \ 2|Blue) = \frac{P(Draw \ from \ 2 \ and \ Blue)}{P(Blue)} = \frac{P(Blue | Draw \ from \ 2) \times P(Draw \ from \ 2)}{P(Blue)}$ = $\frac{\frac{37}{37+23} \times \frac{1}{3}}{0.42}$ = `r round((37/(37+23) * 1/3)/0.42, 2)`

# Problem 4 

```{r}

# Preliminary work with the data 

tab=read.table('Instructor Example Code/Data/PA_HD_1988_35-54.txt',sep='\t',
               header=TRUE,
               stringsAsFactors=FALSE)

# subset just the total rows so that we can compare rural vs urban counties 
tab=tab[tab$Notes=='Total',]

# There are 67 total counties - remove a row that contains total PA data
tab <- tab[!is.na(tab$County.Code), ]

# define flag for analysis 
tab$Urban = ifelse(tab$Population > 20000, "Yes", "No")
     
# In Instructor code Nr = number of races = 2, we have number of ubran/rural classes = 2 
N_classes = length(unique(tab$Urban)) 
# In Instructor code Ns = number of counties = 67, we have the same number
N_counties = nrow(tab)

# state wide average mortality rate 
with(tab, sum(Deaths)/sum(Population)) -> state_wide_rate

# calculate county specific death rate, and evalaute if it higher or lower than the state-wide average 

tab <- 
  tab %>% 
  mutate(
    state_death_rate = Deaths/Population, 
    over_flag = ifelse(state_death_rate > state_wide_rate, 1, 0)
  )

# now we need to get the number of counties by class where we have the rate that is higher than the statewide average 
tab %>% 
  group_by(Urban) %>% 
  summarise(n = n(), 
            outcome = sum(over_flag), 
            outcome_rate = sum(over_flag)/n()) -> summary_data 

# now using data estimates of N and Outcomes, sample from non-informative prior beta distribution 
ns = summary_data %>% select(n) %>% unlist() # order: Rural, Urban 
ys = summary_data %>% select(outcome) %>% unlist() # same ordering appleis to this vector 

# non-informative prior with parameters a = 1, b = 1
a = 1
b = 1

# these are posterior distributions of proportion or the relative frequency of counties with death rate higher than the state-
# wide mortality rate 

set.seed(1)
theta_rural = rbeta(10000, shape1 =  ys[1] + a, # this one is ALPHA parameter 
                    shape2 = ns[1] - ys[1] + b # this one is BETA 
                      )

theta_urban = rbeta(10000, shape1 =  ys[2] + a, # this one is ALPHA parameter 
                    shape2 = ns[2] - ys[2] + b # this one is BETA 
                      )

# take the difference and plot the distribution 
```

### 4 - A

Among 10000 draws from estimated posterior distributions, there were `r length(which(theta_urban - theta_rural > 0))` cases 
where the relative frequency of counties with higher than average mortality rate in the Urban class was higher than the 
relative frequency of counties in the Rural class. This corresponds to the `r round(length(which(theta_urban - theta_rural > 0))/10000, 4)*100`% chance that the true rate of counties with higher than average mortality in the urban class is above that or 
counties in the rural class. At the $\alpha = 0.05$ statistical significance level, we can conclude that the the rate of deaths in
urban counties must be consistently lower

```{r}

# differnece in draws from posterior distributions 
difference = theta_urban - theta_rural

ggplot(data = data.frame(d = difference), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 

```

### 4 - B 

```{r}
# data preliminary, need to work with the rate of death per 1,000
# here we treat counties as data points, aggregate # deaths to a class levels and get a mortality rate, 
# then we take mortality rate times 1,000 to get mortality rate per 1,000 people 

tab %>% 
  group_by(Urban) %>% 
  summarise(n = sum(Population), 
            y = sum(Deaths)) -> summary_data 

ys = summary_data %>% select(y) %>% unlist()
ns = summary_data %>% select(n) %>% unlist()

n = 1000
data_driven_mortality_rate_1000 = state_wide_rate * n 

# common to sample means of poisson (or rates ) from gamma distribution, so that is what we do 
# we obtain posterior lambda by taking data count of death within a county class plus prior rate as shape (ALPHA) parameter
#                                     and number of people in the county class plus prior n as rate (BETA) parameter

# we can also treat prior N as prior OFFSET Ns from data will be OFFSET's obtained from the data 

set.seed(2)
lambda_rural=rgamma(10000,ys[1]+data_driven_mortality_rate_1000,ns[1]+n)
lambda_urban=rgamma(10000,ys[2]+data_driven_mortality_rate_1000,ns[2]+n)

differ=lambda_urban-lambda_rural

```

Among 10000 draws from estimated posterior distributions, there were `r length(which(lambda_urban - lambda_rural > 0))` cases 
where the death rate per 1,000 people in the urban counties was higher than the death rate per 1,000 in the rural counties. 
This corresponds to the `r round(length(which(lambda_urban - lambda_rural > 0))/10000, 4)*100`% chance that the true rate of deaths per 1,000 is higher in the urban counties is higher than that of rural counties. 

While this chance is higher than the traditional 5% frequentist cut off rate for the significance level, estimated chance is 
still fairly slim. 

```{r}
ggplot(data = data.frame(d = differ), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 
```

\newpage 

# Appendix 

**Increasing A, B to 12, 12** for 4-A

```{r}

# non-informative prior with parameters a = 1, b = 1
set.seed(12)
a = 12
b = 12
# Preliminary work with the data 

tab=read.table('Instructor Example Code/Data/PA_HD_1988_35-54.txt',sep='\t',
               header=TRUE,
               stringsAsFactors=FALSE)

# subset just the total rows so that we can compare rural vs urban counties 
tab=tab[tab$Notes=='Total',]

# There are 67 total counties - remove a row that contains total PA data
tab <- tab[!is.na(tab$County.Code), ]

# define flag for analysis 
tab$Urban = ifelse(tab$Population > 20000, "Yes", "No")
     
# In Instructor code Nr = number of races = 2, we have number of ubran/rural classes = 2 
N_classes = length(unique(tab$Urban)) 
# In Instructor code Ns = number of counties = 67, we have the same number
N_counties = nrow(tab)

# state wide average mortality rate 
with(tab, sum(Deaths)/sum(Population)) -> state_wide_rate

# calculate county specific death rate, and evalaute if it higher or lower than the state-wide average 

tab <- 
  tab %>% 
  mutate(
    state_death_rate = Deaths/Population, 
    over_flag = ifelse(state_death_rate > state_wide_rate, 1, 0)
  )

# now we need to get the number of counties by class where we have the rate that is higher than the statewide average 
tab %>% 
  group_by(Urban) %>% 
  summarise(n = n(), 
            outcome = sum(over_flag), 
            outcome_rate = sum(over_flag)/n()) -> summary_data 

# now using data estimates of N and Outcomes, sample from non-informative prior beta distribution 
ns = summary_data %>% select(n) %>% unlist() # order: Rural, Urban 
ys = summary_data %>% select(outcome) %>% unlist() # same ordering appleis to this vector 

# these are posterior distributions of proportion or the relative frequency of counties with death rate higher than the state-
# wide mortality rate 
theta_rural = rbeta(10000, shape1 =  ys[1] + a, # this one is ALPHA parameter 
                    shape2 = ns[1] - ys[1] + b # this one is BETA 
                      )

theta_urban = rbeta(10000, shape1 =  ys[2] + a, # this one is ALPHA parameter 
                    shape2 = ns[2] - ys[2] + b # this one is BETA 
                      )
difference = theta_urban - theta_rural

ggplot(data = data.frame(d = difference), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 
# take the difference and plot the distribution 
```

* New chance: `r round(length(which(theta_urban - theta_rural > 0))/10000, 4)*100`%

* Estimated Difference: `r mean(theta_urban - theta_rural) %>% round(., 2)`

* Estimated SD of distribution: `r sd(theta_urban - theta_rural) %>% round(., 2)`

**Increasing A, B to 25, 25** for 4-A
```{r}

# non-informative prior with parameters a = 1, b = 1
set.seed(123)
a = 25
b = 25
# Preliminary work with the data 

tab=read.table('Instructor Example Code/Data/PA_HD_1988_35-54.txt',sep='\t',
               header=TRUE,
               stringsAsFactors=FALSE)

# subset just the total rows so that we can compare rural vs urban counties 
tab=tab[tab$Notes=='Total',]

# There are 67 total counties - remove a row that contains total PA data
tab <- tab[!is.na(tab$County.Code), ]

# define flag for analysis 
tab$Urban = ifelse(tab$Population > 20000, "Yes", "No")
     
# In Instructor code Nr = number of races = 2, we have number of ubran/rural classes = 2 
N_classes = length(unique(tab$Urban)) 
# In Instructor code Ns = number of counties = 67, we have the same number
N_counties = nrow(tab)

# state wide average mortality rate 
with(tab, sum(Deaths)/sum(Population)) -> state_wide_rate

# calculate county specific death rate, and evalaute if it higher or lower than the state-wide average 

tab <- 
  tab %>% 
  mutate(
    state_death_rate = Deaths/Population, 
    over_flag = ifelse(state_death_rate > state_wide_rate, 1, 0)
  )

# now we need to get the number of counties by class where we have the rate that is higher than the statewide average 
tab %>% 
  group_by(Urban) %>% 
  summarise(n = n(), 
            outcome = sum(over_flag), 
            outcome_rate = sum(over_flag)/n()) -> summary_data 

# now using data estimates of N and Outcomes, sample from non-informative prior beta distribution 
ns = summary_data %>% select(n) %>% unlist() # order: Rural, Urban 
ys = summary_data %>% select(outcome) %>% unlist() # same ordering appleis to this vector 

# these are posterior distributions of proportion or the relative frequency of counties with death rate higher than the state-
# wide mortality rate 
theta_rural = rbeta(10000, shape1 =  ys[1] + a, # this one is ALPHA parameter 
                    shape2 = ns[1] - ys[1] + b # this one is BETA 
                      )

theta_urban = rbeta(10000, shape1 =  ys[2] + a, # this one is ALPHA parameter 
                    shape2 = ns[2] - ys[2] + b # this one is BETA 
                      )
difference = theta_urban - theta_rural

ggplot(data = data.frame(d = difference), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 
# take the difference and plot the distribution 
```


* New chance: `r round(length(which(theta_urban - theta_rural > 0))/10000, 4)*100`%

* Estimated Difference: `r mean(theta_urban - theta_rural) %>% round(., 2)`

* Estimated SD of distribution: `r sd(theta_urban - theta_rural) %>% round(., 2)`

**Increasing A, B to 100, 100** for 4-A
```{r}

# non-informative prior with parameters a = 1, b = 1
set.seed(123)
a = 100
b = 100
# Preliminary work with the data 

tab=read.table('Instructor Example Code/Data/PA_HD_1988_35-54.txt',sep='\t',
               header=TRUE,
               stringsAsFactors=FALSE)

# subset just the total rows so that we can compare rural vs urban counties 
tab=tab[tab$Notes=='Total',]

# There are 67 total counties - remove a row that contains total PA data
tab <- tab[!is.na(tab$County.Code), ]

# define flag for analysis 
tab$Urban = ifelse(tab$Population > 20000, "Yes", "No")
     
# In Instructor code Nr = number of races = 2, we have number of ubran/rural classes = 2 
N_classes = length(unique(tab$Urban)) 
# In Instructor code Ns = number of counties = 67, we have the same number
N_counties = nrow(tab)

# state wide average mortality rate 
with(tab, sum(Deaths)/sum(Population)) -> state_wide_rate

# calculate county specific death rate, and evalaute if it higher or lower than the state-wide average 

tab <- 
  tab %>% 
  mutate(
    state_death_rate = Deaths/Population, 
    over_flag = ifelse(state_death_rate > state_wide_rate, 1, 0)
  )

# now we need to get the number of counties by class where we have the rate that is higher than the statewide average 
tab %>% 
  group_by(Urban) %>% 
  summarise(n = n(), 
            outcome = sum(over_flag), 
            outcome_rate = sum(over_flag)/n()) -> summary_data 

# now using data estimates of N and Outcomes, sample from non-informative prior beta distribution 
ns = summary_data %>% select(n) %>% unlist() # order: Rural, Urban 
ys = summary_data %>% select(outcome) %>% unlist() # same ordering appleis to this vector 

# these are posterior distributions of proportion or the relative frequency of counties with death rate higher than the state-
# wide mortality rate 
theta_rural = rbeta(10000, shape1 =  ys[1] + a, # this one is ALPHA parameter 
                    shape2 = ns[1] - ys[1] + b # this one is BETA 
                      )

theta_urban = rbeta(10000, shape1 =  ys[2] + a, # this one is ALPHA parameter 
                    shape2 = ns[2] - ys[2] + b # this one is BETA 
                      )
difference = theta_urban - theta_rural

ggplot(data = data.frame(d = difference), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 
# take the difference and plot the distribution 
```


* New chance: `r round(length(which(theta_urban - theta_rural > 0))/10000, 4)*100`%

* Estimated Difference: `r mean(theta_urban - theta_rural) %>% round(., 2)`

* Estimated SD of distribution: `r sd(theta_urban - theta_rural) %>% round(., 2)`


**Increasing A, B to 200, 200** for 4-A
```{r}

# non-informative prior with parameters a = 1, b = 1
set.seed(123)
a = 200
b = 200
# Preliminary work with the data 

tab=read.table('Instructor Example Code/Data/PA_HD_1988_35-54.txt',sep='\t',
               header=TRUE,
               stringsAsFactors=FALSE)

# subset just the total rows so that we can compare rural vs urban counties 
tab=tab[tab$Notes=='Total',]

# There are 67 total counties - remove a row that contains total PA data
tab <- tab[!is.na(tab$County.Code), ]

# define flag for analysis 
tab$Urban = ifelse(tab$Population > 20000, "Yes", "No")
     
# In Instructor code Nr = number of races = 2, we have number of ubran/rural classes = 2 
N_classes = length(unique(tab$Urban)) 
# In Instructor code Ns = number of counties = 67, we have the same number
N_counties = nrow(tab)

# state wide average mortality rate 
with(tab, sum(Deaths)/sum(Population)) -> state_wide_rate

# calculate county specific death rate, and evalaute if it higher or lower than the state-wide average 

tab <- 
  tab %>% 
  mutate(
    state_death_rate = Deaths/Population, 
    over_flag = ifelse(state_death_rate > state_wide_rate, 1, 0)
  )

# now we need to get the number of counties by class where we have the rate that is higher than the statewide average 
tab %>% 
  group_by(Urban) %>% 
  summarise(n = n(), 
            outcome = sum(over_flag), 
            outcome_rate = sum(over_flag)/n()) -> summary_data 

# now using data estimates of N and Outcomes, sample from non-informative prior beta distribution 
ns = summary_data %>% select(n) %>% unlist() # order: Rural, Urban 
ys = summary_data %>% select(outcome) %>% unlist() # same ordering appleis to this vector 

# these are posterior distributions of proportion or the relative frequency of counties with death rate higher than the state-
# wide mortality rate 
theta_rural = rbeta(10000, shape1 =  ys[1] + a, # this one is ALPHA parameter 
                    shape2 = ns[1] - ys[1] + b # this one is BETA 
                      )

theta_urban = rbeta(10000, shape1 =  ys[2] + a, # this one is ALPHA parameter 
                    shape2 = ns[2] - ys[2] + b # this one is BETA 
                      )
difference = theta_urban - theta_rural

ggplot(data = data.frame(d = difference), 
       aes(x = d)) + 
  theme_minimal() + 
  geom_histogram(color = 'black', fill = 'lightgrey', bins = 30) + 
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") + 
  labs(title = "Distribution of differences in Urban vs Rural \ncounty relative frequencies of higher death rates", 
       x = "Urban minus Rural Difference", 
       y = "Count") 
# take the difference and plot the distribution 
```


* New chance: `r round(length(which(theta_urban - theta_rural > 0))/10000, 4)*100`%

* Estimated Difference: `r mean(theta_urban - theta_rural) %>% round(., 2)`

* Estimated SD of distribution: `r sd(theta_urban - theta_rural) %>% round(., 2)`